---
title: "DataPrep"
format: pdf
author: 'Jincheng Li, 
         Mason Spikula, 
         Quinn Hughes'
---

```{r load libraries}
library(readr)
library(dplyr)
library(purrr)
library(ggplot2)
```

### Loading Datasets

```{r load datasets}
cell_per_100 <- read_csv("cell_phones_per_100_people.csv")
gap <- read_csv('gapminder_2025.csv')
hdi <- read_csv('hdi_human_development_index.csv')
maternal_mortality <- read_csv('maternal_mortality_ratio_who.csv')
daily_inc <- read_csv('mincpcap_cppp (1).csv')
engy_use_pctg <- read_csv('residential_energy_use_percent.csv')
```

### Pulling Data From Year We Want and Cleaning it

```{r construct panel data}
cell_2010 <- cell_per_100[, c('geo', '2010')]
engy_2010 <- engy_use_pctg[, c('geo', '2010')]
landlocked <- gap[, c('geo', 'landlocked')]
hdi_2010 <- hdi[, c('geo', '2010')]
mat_mort_2010 <- maternal_mortality[, c('geo', '2010')]
daily_inc_2010 <- daily_inc[, c('geo', '2010')]


full_2010 <- list(cell_2010, engy_2010, landlocked, hdi_2010, mat_mort_2010, daily_inc_2010) |>
  reduce(full_join, by = "geo")

clean_2010 <- full_2010 |> 
  na.omit()

# names(clean_2010)

names(clean_2010) <- c('geo', 
                       'cell_phones_per_100_ppl', 
                       'residential_energy_use_pctg', 
                       'landlocked',
                       'human_development_ind', 
                       'maternal_mortality', 
                       'daily_income')

head(clean_2010)
nrow(clean_2010)
glimpse(clean_2010)
summary(clean_2010)
```

### Saving Data Into New File

```{r save cleaned data}
write.csv(clean_2010, "clean_2010.csv", row.names = FALSE)
```

### Displaying Each Variable

#### Landlocked

```{r}
table(clean_2010$landlocked)

ggplot(clean_2010, aes(x = landlocked)) +
  geom_bar() +
  labs(title = "Landlocked vs. Coastline Countries (2010)",
       x = "Landlocked", y = "Number of Countries")
```

#### Daily Income

```{r}
ggplot(clean_2010, aes(x = daily_income)) +
  geom_histogram(bins = 30) +
  labs(title = "Daily Income (2010)", x = "Daily Income", y = "Count")

clean_2010 |>
  summarise(mean   = mean(daily_income),
            median = median(daily_income),
            sd     = sd(daily_income),
            min    = min(daily_income),
            max    = max(daily_income))

clean_2010 |> arrange(desc(daily_income)) |> select(geo, daily_income) |> head(5)
clean_2010 |> arrange(daily_income) |> select(geo, daily_income) |> head(5)
```

#### Cell Phones per 100 People

```{r}
ggplot(clean_2010, aes(x = cell_phones_per_100_ppl)) +
  geom_histogram(bins = 30) +
  labs(title = "Cell Phones per 100 People (2010)", x = "Cell Phones", y = "Count")

clean_2010 |>
  summarise(mean   = mean(cell_phones_per_100_ppl),
            median = median(cell_phones_per_100_ppl),
            sd     = sd(cell_phones_per_100_ppl),
            min    = min(cell_phones_per_100_ppl),
            max    = max(cell_phones_per_100_ppl))

clean_2010 |> arrange(desc(cell_phones_per_100_ppl)) |> select(geo, cell_phones_per_100_ppl) |> head(5)
clean_2010 |> arrange(cell_phones_per_100_ppl) |> select(geo, cell_phones_per_100_ppl) |> head(5)
```

#### Residential Energy Use Percentage

```{r}
ggplot(clean_2010, aes(x = residential_energy_use_pctg)) +
  geom_histogram(bins = 30) +
  labs(title = "Residential Energy Use Percentage (2010)", x = "Residential Energy Use", y = "Count")

clean_2010 |>
  summarise(mean   = mean(residential_energy_use_pctg),
            median = median(residential_energy_use_pctg),
            sd     = sd(residential_energy_use_pctg),
            min    = min(residential_energy_use_pctg),
            max    = max(residential_energy_use_pctg))

clean_2010 |> arrange(desc(residential_energy_use_pctg)) |> select(geo, residential_energy_use_pctg) |> head(5)
clean_2010 |> arrange(residential_energy_use_pctg) |> select(geo, residential_energy_use_pctg) |> head(5)
```

#### Maternal Mortality

```{r}
ggplot(clean_2010, aes(x = maternal_mortality)) +
  geom_histogram(bins = 30) +
  labs(title = "Maternal Mortality (2010)", x = "Maternal Deaths per 100,000 Births", y = "Count")

clean_2010 |>
  summarise(mean   = mean(maternal_mortality),
            median = median(maternal_mortality),
            sd     = sd(maternal_mortality),
            min    = min(maternal_mortality),
            max    = max(maternal_mortality))

clean_2010 |> arrange(desc(maternal_mortality)) |> select(geo, maternal_mortality) |> head(5)
clean_2010 |> arrange(maternal_mortality) |> select(geo, maternal_mortality) |> head(5)
```

#### HDI

There aren't any outliers with HDI because it is a metric that takes multiple factors into account and is on a scale of 0-1

```{r}
ggplot(clean_2010, aes(x = human_development_ind)) +
  geom_histogram(bins = 30) +
  labs(title = "Human Development Index (2010)", x = "HDI", y = "Count")

clean_2010 |>
  summarise(mean   = mean(human_development_ind),
            median = median(human_development_ind),
            sd     = sd(human_development_ind),
            min    = min(human_development_ind),
            max    = max(human_development_ind))
```

### Displaying Graphs of Each Variable Related to One Another

```{r EDA}
pairs(clean_2010[c(
  "cell_phones_per_100_ppl",
  "residential_energy_use_pctg",
  "human_development_ind",
  "maternal_mortality",
  "daily_income"
)])
```

We can expect a high VIF among variables since they are all kind of intercorrelated in some way.

### Creating Log of Data

```{r transform data}
transformed_2010 <- transform(
  clean_2010,
  log_daily_income = log(daily_income),
  log_energy = log1p(residential_energy_use_pctg),
  log_mortality = log1p(maternal_mortality),
  landlocked = factor(landlocked)
)

pairs(transformed_2010[c(
  "cell_phones_per_100_ppl",
  "log_energy",
  "human_development_ind",
  "log_mortality",
  "log_daily_income"
)])

summary(transformed_2010)

```

### Making Lines of Regression

#### Regression of Log(Daily Income) vs Everything

```{r}
income_model1 <- lm(
  log_daily_income ~ cell_phones_per_100_ppl +
    log_energy +
    human_development_ind +
    log_mortality +
    landlocked,
  data = transformed_2010
)

summary(income_model1)

par(mfrow = c(2, 2))
plot(income_model1)
par(mfrow = c(1, 1))
```

#### Regression of Daily Income vs Landlocked and Energy

```{r}
income_model2 <- lm(
  log_daily_income ~ log_energy +
    landlocked,
  data = transformed_2010
)

summary(income_model2)

par(mfrow = c(2, 2))
plot(income_model2)
par(mfrow = c(1, 1))
```

#### Regression Daily Income vs HDI

```{r}
income_model3 <- lm(
  log_daily_income ~ human_development_ind,
  data = transformed_2010
)

summary(income_model3)

par(mfrow = c(2, 2))
plot(income_model3)
par(mfrow = c(1, 1))
```
